Ultra-deep $K_s$-band Imaging of the \textit{Hubble} Frontier Fields

### 2016-04-17 Paper0/ms.tex
small edit
Danilo's edits

During the last two decades, near-infrared (NIR) imaging has gained a dominant role in studies of galaxy formation and evolution.

Detection of galaxies in the $K$-band ($\lambda_{\rm eff} \sim 2.2$~$\mu$m) provided the first opportunity to construct a comprehensive picture of the population of galaxies in the early universe, as it enabled the discovery of galaxies at $z > 2$ with suppressed emission in the optical (i.e., rest-frame UV) due to evolved stellar populations and/or significant amount of dust extinction (e.g., \textcolor{red}{refs}). In fact, these galaxies, which dominate the high-mass end of the high-$z$ galaxy population, were previously missed by rest-frame UV selection techniques (e.g., $U$-dropout galaxies; \textcolor{red}{refs}). Imaging in the $K$-band allows for the direct probing out to $z\approx5$ of rest-frame wavelengths longer than the Balmer break. Sampling the rest-frame optical wavelength regime is crucial for high-$z$ studies, as significantly less affected by dust obscuration and a better probe of the galaxy stellar mass compared to the rest-frame UV (more sensitive to unobscured star formation). While imaging at even longer wavelengths ($\lambda>3$~$\mu$m) are needed to probe the rest-frame optical of galaxies in the first billion years of cosmic history, deep sub-arsec resolution imaging at these wavelengths will not be available until the advent of the {\it James Webb Space Telescope} ({\it JWST}). Deep $K$-band data with good enough (i.e., FWHM$<1^{\prime \prime}$) image quality have been fundamental to fully exploit {\it Spitzer}-IRAC imaging data characterized by much poorer spatial resolution, allowing for the correction of the resulting blending of sources in IRAC images. Finally, for space-based studies, $K$ band crucially fills the gap between the reddest {\it Hubble Space Telescope} ({\it HST}) filter (i.e., F160W, with $\lambda \sim 1.6$~$\mu$m) and the IRAC 3.6~$\mu$m pass-band, greatly improving the constraints on both the redshifts and the stellar-population properties of galaxies.

It is not surprising that, given the aforementioned reasons, all successful extragalactic surveys performed in the last fifteen years to investigate galaxies populations in the early universe invested significant resources to obtain deep NIR imaging data with good image quality, especially in the $K$ band, usually adopted as the detection band. Indeed, our understanding of galaxy formation and evolution in the last couple of decades has impressively proceeded forward mostly driven by NIR progressively deeper, wider, and with better image quality, surveys, such as ...  FIRES, FIREWORKS, NMBS, UltraVISTA, etc... Subaru Super Deep Field (AO) \cite{minowa:05}... HUGS \citep{hugs}.\textcolor{red}{-- we should provide a complete list of NIR surveys here -- }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                                                %
%  Summary table
%                                                                %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{deluxetable*}{ccccrccccc}
%\tabletypesize{\footnotesize}
\tablewidth{0pt} 
\tablecaption{Field summary \label{tab:summary}} 
\tablehead{ 
\colhead{Field} & \colhead{R.A.} & \colhead{Dec.} & \colhead{Instrument} & \colhead{Epoch} & \textit{HST}\tablenotemark{a} & \colhead{$t$, hours} & {Depth\tablenotemark{b}} & \colhead{FWHM}}
\startdata 
Abell 2744   & $00:14:21.2$ & $-30:23:50$ & \textit{VLT}/\HAWKI     & 2013 Oct 24--2013 Dec 24 & C+P & 29.3 &  26.0 & $0\farcs39$ \\ 
MACS-0416    & $04:16:08.9$ & $-24:04:28$ & \textit{VLT}/\HAWKI     & 2013 Oct 25--2014 Feb 23 & C+P & 25.8 &  26.0 & $0\farcs36$ \\ 
Abell S1063  & $22:49:01.1$ & $-44:32:13$ & \textit{VLT}/\HAWKI     & 2015 Jul 8--Sep 22       & C+P & 27.9 &  26.0 & $0\farcs39$ \\ 
Abell 370    & $02:40:03.3$ & $-01:36:23$ & \textit{VLT}/\HAWKI     & 2015 Jul 26--2016 Jan 28 & C+P & 28.3 &  26.0 & $0\farcs35$ \\ 
\vspace {-0.2cm}\\ 
\multirow{2}{*}{MACS-0717}    & \multirow{2}{*}{$07:17:34.0$} & \multirow{2}{*}{$+37:44:49$} &  \multirow{2}{*}{\textit{Keck}/MOSFIRE}  &  \multirow{2}{*}{2015 Jan 26--27, 2016 Jan 21--22}  & C & 4.3 & 25.3 & $0\farcs42$ \\ 
                              &                             &                             &                                          &      & P & 3.8 & 25.5 & $0\farcs49$ \\ 
\vspace {-0.2cm}\\ 
\multirow{4}{*}{MACS-1149}    & \multirow{4}{*}{$11:49:36.3$} & \multirow{4}{*}{$+22:23:58$} & \multirow{2}{*}{\textit{Keck}/MOSFIRE}  & \multirow{2}{*}{2015 Feb 24, 2016 Jan 21--22}      & C & 5.5 & 25.2 & $0\farcs53$ \\
&  &         &      &   & P & 4.8 & 25.1 & $0\farcs54$ \\
\vspace {-0.4cm}\\ 
&  &         & \textit{VLT}/\HAWKI\tablenotemark{c}     & 2013 Mar 21--2014 Jun 9      & C & 5.3 & 25.0 & $0\farcs41$
\enddata 
%\vspace{-0.8cm} 
\tablenotetext{a}{Coverage of \textit{Hubble} survey fields: C=Cluster, P=Parallel.  Most \HAWKI\ pointings cover the two \textit{HST} fields simultaneously (C+P; Fig.~\ref{fig:layout}). The MOSFIRE observations of MACS-0717 and MACS-1149 require two separate pointings of the instrument to cover the cluster and parallel \textit{HST} fields, which have the different characteristics as indicated.}
%\footnotetext[2]{Approximate exposure time equivalent at $0\farcs4$ seeing:  $t_{0.4}=\sum_i t_i \cdot (0.4/\mathrm{FWHM}_i)^2$.}
\tablenotetext{b}{Depth is defined as the 5$\sigma$ limiting magnitude for point sources, measured in $D=0\farcs6$ apertures (\S\ref{s:depth}).} 
\tablenotetext{c}{Archival observations from ESO program \href{http://archive.eso.org/wdb/wdb/eso/eso_archive_main/query?prog_id=090.A-0458(A)}{090.A-0458} (PI: Infante).}
\end{deluxetable*}
%\vspace{1cm}

The latest effort to further our knowledge of galaxy formation and evolution is represented by the HST Frontier Fields program. \textcolor{red}{-- brief description of HFF, followed by the need for mega-deep K-band data --}

@@

$K_s$-band observations of the Abell 2744 and MACS-0416 fields were obtained between 2013 Oct and 2014 Feb with the High Acuity Wide-field K-band Imager \citep[\HAWKI;][]{hawki, hawki2} on the 8.2\,m UT4 telescope at the ESO \textit{Very Large Telescope} (ESO program \mbox{092.A-0472}).  \HAWKI\ $K_s$-badn bservations of the Abell S1063 and Abell 370 fields were obtained between 2015 Jul and 2016 Jan from a subsequent program (095.A-0533).  \HAWKI\ is composed of four chips with 2048$\times$2048 $0\farcs106$ pixels, and the full $7\farcm5\times7\farcm5$ \HAWKI\ field-of-view is perfectly suited to cover both the primary and parallel \textit{HST} ACS optical and WFC3 IR fields simultaneously in a single pointing (see Fig.\ref{fig:layout}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   Exposure Map figure
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
\epsscale{1.18}
\plotone{hawkiff_layout.pdf} % Figures/exposure_map.pdf
\caption{Layout of the Frontier Fields $K_s$-band mosaics.  The positions of the \textit{HST} cluster and parallel fields are shown in the blue (ACS optical) and red (WFC3 IR) polygons.  The light blue polygons in the MACS-0717 field show additional wide-field ACS imaging coverage from  programs GO-9722 and GO-10420 \citep[PI: Ebeling;][]{ma:11}.  The area covered by deep imaging in the \textit{Spitzer}/IRAC 3.6 and 4.5\,\mum\ channels is shown in orange, with the cluster fields indicated by the dashed lines.  The \HAWKI\ field of view is perfectly suited (Abell 2744, MACS-0416, Abell S1063 and Abell 370) for simultaneous imaging of the cluster+parallel field pairs, which require two separate pointings with MOSFIRE (MACS-0717 and MACS-1149).  The footprint of additional archival \HAWKI\ coverage of the MACS-1149 field is indicated by the gray square in the lower right panel.\label{fig:layout}}  
\end{figure*}
%%%%%%%%%%%%%%%%

@@

The \HAWKI\ and MOSFIRE observations were reduced with a pipeline that has been developed for previous surveys with the NEWFIRM \citep[NMBS;][]{whitaker:nmbs} and FOURSTAR \citep[zFOURGE;][, Labb\'e in prep]{spitler:12} infrared imaging instruments.  Treating each detector individually, the pipeline is easily modified for the different instrument configurations of the four \HAWKI\ and single MOSFIRE chips.  The primary task of the pipeline is removing the bright, time-variable sky background from the individual exposures, which is often some $10^4$ times brighter than the distant galaxies of interest in the field.  With such a bright background, we first determine an empirical ``sky flat'' that is a median of all of the \textit{science} exposures in a \HAWKI\ OB or MOSFIRE group, after rejecting the brightest 12 exposures at each pixel position to remove the contribution of bright objects.  We find these empirical flats to be preferable to external twilight or dome flats given the difficulty of obtaining a truly flat illumination pattern over such large detector fields-of-view.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   2MASS comparison
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
\epsscale{1.18}
\plotone{2mass_phot.pdf} % Figures/exposure_map.pdf
\caption{Comparison of bright-star photometry to the 2MASS public catalog.  The photometric calibration of the $K_s$-band images were determined from standard star observations taken concurrently with the science exposures.  The uncertainties shown are taken from the 2MASS catalog; these stars are measured at very high S/N in the deep $K_s$-band mosaics.  We estimate the overall systematic uncertainty in the photometric calibration to be $\lesssim0.05~\mathrm{mag}$.\label{fig:2mass}}  
\end{figure}
%%%%%%%%%%%%%

@@

The 2-Micron All Sky Survey \citep[2MASS;][]{2mass} provides an additional check on the photometric calibration, though the comparison is limited due to relatively little brightness overlap between the faint end of reliable 2MASS photometry and the bright end where stars are in the linear regime of the detectors on the 8--10\,m telescopes.  A comparison of the observed photometry in the four survey fields to the 2MASS catalog magnitudes is shown in Fig.~\ref{fig:2mass}.  The stellar photometry on the deep $K_s$-band mosaics is measured within $1\arcsec$ apertures corrected to infinity with the curves of growth described below.  The error-bars on the points in Fig.~\ref{fig:2mass} come from the 2MASS catalog; the photometric uncertainties are negligible for such bright stars in the deep images ($\mathrm{S/N}>100$).  The 2MASS comparison suggests a systematic offset of $-0.09$~mag for the Abell S1063 field.  However, the exposure time for that field was similar to those of the additional deep \HAWKI\ fields and the derived depths of \textit{all} deep \HAWKI\ fields are nearly identical (Table~\ref{tab:summary}).  Given the overall agreement with the 2MASS photometry, we estimate that the systematic uncertainty of the absolute photometric calibration is $\sigma_\mathrm{sys}\leq0.05$~mag.

@@

In order to simplify measurements from the $K_s$-band images in consort with space-based \textit{HST} and \textit{Spitzer} mosaics, the astrometry of the ground-based mosaics must be refined.  Reference absolute astrometric catalogs were generated from \textit{HST} images when available (i.e., the chips overlapping the cluster and parallel \textit{HST} fields) and public \textit{Subaru} Suprime-Cam $r_c$-band images otherwise\footnote{\url{http://archive.stsci.edu/pub/hlsp/clash/}}.  Next, object catalogs are generated with the SEXtractor software \citep{bertin:96} for each individual background-subtracted exposure and transformations to the reference frame (shift, rotation and scale) are computed with with the \texttt{stsci.stimage.xyxymatch} and \texttt{scikit-image.transform}\footnote{\url{http://scikit-image.org}} Python software tools.  For the \HAWKI\ exposures, we fit a third-order polynomial to the geometric distortion model determined by \cite{libralato:14} and specify the polynomial terms as ``Simple Imaging Polynomial'' (SIP) coefficients \citep{fits:sip} in the individual exposure FITS files.  The SIP FITS header generally applicable for \HAWKI\ is provided in Appendix \xxx.  We note that the \HAWKI\ distortion is generally small, reaching $\sim$2 pixels ($0\farcs2$) at the image corners, but that the distortion corrections are necessary, in particular, given the excellent overall image quality of the observations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   Drizzle figure
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!ht]
\epsscale{1.18}
\plotone{ivar_mosaics.pdf} % Figures/exposure_map.pdf
\caption{Inverse variance maps of the $K_s$-band mosaics.  The orientation of the fields is the same as that shown in Fig.~\ref{fig:layout}.  The four \HAWKI\ chips and two MOSFIRE pointings are clearly visible for the first four and last two fields, respectively.
\label{fig:ivar_mosaics}}  
\end{figure}
%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   Image quality figure
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[!t]
\epsscale{1.18}
\plotone{curve_of_growth.pdf} % Figures/exposure_map.pdf
\caption{Curves of growth of stellar profiles in the Frontier Fields $K_s$-band mosaics.  The inverse aperture corrections for point sources in $D=0\farcs6$ apertures are indicated in each panel.  The profiles of stars are well-fit by Moffat profiles with $\mathrm{FWHM}\sim0\farcs4$ in most cases and $\beta\sim2.1$.  These extended Moffat profiles appear to be characteristic of deep $K_s$-band images and they have substantially more flux at large radii ($r>\mathrm{FWHM}/2$) than Gaussian profiles with the same FWHM (orange curves in the left panels).  \label{fig:cog}}  
\end{figure*}
%%%%%%%%%%%%%%

@@

Finally, we subtract a cell-based background from the final mosaics using an algorithm based on that used by SExtractor and SWarp but that provides more aggressive masking of flux in the outer isophotes of bright galaxies. The mosaics are shown in Fig.~\ref{fig:layout}, along with the position of the deep Frontier Fields \textit{HST} imaging fields.  We also compute the robust NMAD scatter \citep{brammer:08} of empty sky pixels within the same cells to empirically calibrate the inverse variance maps that are shown in Fig.~\ref{fig:ivar_mosaics} (see also \S\ref{s:depth}).  The final image and inverse variance mosaics are provided as an ESO Phase 3 data product \xxx\ URL, as well as attached as a digital product to this manuscript (\xxx\ I think ApJSS can and wants to do this).  The released images are all scaled to a common zeropoint of 26.0 (AB mag), which gives pixel values of order unity faint galaxies near the detection threshold.  In the sections below we describe the characteristics of the mosaics in more detail.

>>>>

During the last two decades, near-infrared (NIR) imaging has gained a dominant role in studies of galaxy formation and evolution, enabling transformational advances in our understanding of galaxy populations at early cosmic times.

Detection of galaxies in the $K$-band ($\lambda_{\rm eff} \sim 2.2$~$\mu$m) provided the first opportunity to construct a comprehensive picture of the population of galaxies in the early universe, as it enabled the discovery of galaxies at $z > 2$ with suppressed emission in the optical (i.e., rest-frame UV) due to evolved stellar populations and/or significant amount of dust extinction (e.g., \textcolor{red}{refs}). In fact, these galaxies, which dominate the high-mass end of the high-$z$ galaxy population, were previously missed by rest-frame UV selection techniques (e.g., $U$-dropout galaxies; \textcolor{red}{refs}). Imaging in the $K$-band allows for the direct probing out to $z\approx5$ of rest-frame wavelengths longer than the Balmer break. Sampling the rest-frame optical wavelength regime is crucial for high-$z$ studies, as significantly less affected by dust obscuration and a better probe of the galaxy stellar mass compared to the rest-frame UV (more sensitive to unobscured star formation). While imaging at even longer wavelengths ($\lambda>3$~$\mu$m) are needed to probe the rest-frame optical of galaxies in the first billion years of cosmic history, deep sub-arsec resolution imaging at these wavelengths will not be available until the advent of the {\it James Webb Space Telescope} ({\it JWST}). Deep $K$-band data with good enough (i.e., FWHM$<1^{\prime \prime}$) image quality have been fundamental to fully exploit {\it Spitzer}-IRAC imaging data characterized by much poorer spatial resolution, allowing for the mitigation of the resulting blending of sources in IRAC images. Finally, for space-based studies, $K$ band crucially fills the gap between the reddest {\it Hubble Space Telescope} ({\it HST}) filter (i.e., F160W, with $\lambda \sim 1.6$~$\mu$m) and the IRAC 3.6~$\mu$m pass-band, greatly improving the constraints on both the redshifts and the stellar-population properties of galaxies.

It is not surprising that, given the aforementioned reasons, all successful extragalactic surveys performed in the last fifteen years to investigate galaxies populations in the early universe invested significant resources to obtain deep NIR imaging data with good image quality, especially in the $K$ band, usually adopted as the detection band. Indeed, our understanding of galaxy formation and evolution in the last couple of decades has impressively proceeded forward mostly driven by progressively deeper, wider, and with better image quality, NIR surveys, such as ...  FIRES, FIREWORKS, NMBS, UltraVISTA, etc... Subaru Super Deep Field (AO) \cite{minowa:05}... HUGS \citep{hugs}.\textcolor{red}{xxx --  we should provide a complete list of NIR surveys here -- }

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%                                                                %
%  Summary table
%                                                                %
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{deluxetable*}{ccccrccccc}
%\tabletypesize{\footnotesize}
\tablewidth{0pt} 
\tablecaption{Field summary \label{tab:summary}} 
\tablehead{ 
\colhead{Field} & \colhead{R.A.} & \colhead{Dec.} & \colhead{Instrument} & \colhead{Epoch} & \textit{HST}\tablenotemark{a} & \colhead{$t$, hours} & {Depth\tablenotemark{b}} & \colhead{FWHM}}
\startdata 
Abell 2744   & $00:14:21.2$ & $-30:23:50$ & \textit{VLT}/\HAWKI     & 2013 Oct 24--2013 Dec 24 & C+P & 29.3 &  26.0 & $0\farcs39$ \\ 
MACS-0416    & $04:16:08.9$ & $-24:04:28$ & \textit{VLT}/\HAWKI     & 2013 Oct 25--2014 Feb 23 & C+P & 25.8 &  26.0 & $0\farcs36$ \\ 
Abell S1063  & $22:49:01.1$ & $-44:32:13$ & \textit{VLT}/\HAWKI     & 2015 Jul 8--Sep 22       & C+P & 27.9 &  26.0 & $0\farcs39$ \\ 
Abell 370    & $02:40:03.3$ & $-01:36:23$ & \textit{VLT}/\HAWKI     & 2015 Jul 26--2016 Jan 28 & C+P & 28.3 &  26.0 & $0\farcs35$ \\ 
\vspace {-0.2cm}\\ 
\multirow{2}{*}{MACS-0717}    & \multirow{2}{*}{$07:17:34.0$} & \multirow{2}{*}{$+37:44:49$} &  \multirow{2}{*}{\textit{Keck}/MOSFIRE}  &  \multirow{2}{*}{2015 Jan 26--27, 2016 Jan 21--22}  & C & 4.3 & 25.3 & $0\farcs42$ \\ 
                              &                             &                             &                                          &      & P & 3.8 & 25.5 & $0\farcs49$ \\ 
\vspace {-0.2cm}\\ 
\multirow{4}{*}{MACS-1149}    & \multirow{4}{*}{$11:49:36.3$} & \multirow{4}{*}{$+22:23:58$} & \multirow{2}{*}{\textit{Keck}/MOSFIRE}  & \multirow{2}{*}{2015 Feb 24, 2016 Jan 21--22}      & C & 5.5 & 25.2 & $0\farcs53$ \\
&  &         &      &   & P & 4.8 & 25.1 & $0\farcs54$ \\
\vspace {-0.4cm}\\ 
&  &         & \textit{VLT}/\HAWKI\tablenotemark{c}     & 2013 Mar 21--2014 Jun 9      & C & 5.3 & 25.0 & $0\farcs41$
\enddata 
%\vspace{-0.8cm} 
\tablenotetext{a}{Coverage of \textit{Hubble} survey fields: C=Cluster, P=Parallel.  Most \HAWKI\ pointings cover the two \textit{HST} fields simultaneously (C+P; Fig.~\ref{fig:layout}). \\The MOSFIRE observations of MACS-0717 and MACS-1149 require two separate pointings of the instrument to cover the cluster and pa-\\rallel \textit{HST} fields, which have the different characteristics as indicated.}
%\footnotetext[2]{Approximate exposure time equivalent at $0\farcs4$ seeing:  $t_{0.4}=\sum_i t_i \cdot (0.4/\mathrm{FWHM}_i)^2$.}
\tablenotetext{b}{Depth is defined as the 5$\sigma$ limiting magnitude for point sources, measured in $D=0\farcs6$ apertures (\S\ref{s:depth}).} 
\tablenotetext{c}{Archival observations from ESO program \href{http://archive.eso.org/wdb/wdb/eso/eso_archive_main/query?prog_id=090.A-0458(A)}{090.A-0458} (PI: Infante).}
\end{deluxetable*}
%\vspace{1cm}

The latest effort to further our knowledge of galaxy formation and evolution is represented by the {\it HST} Frontier Fields (HFF) program. The HFF program is a multi-cycle program consisting of $\sim$560 orbits of Director's Discretionary (DD) time that is imaging six deep fields centered on strong lensing galaxy clusters in parallel with six deep blank fields. The primary science goals of the twelve HFF fields are to 1) reveal the population of galaxies at $z=5-10$ that are 10-50 times fainter intrinsically than any presently known, 2) solidify our understanding of the stellar masses and star formation histories of faint galaxies, 3) provide the first statistically meaningful morphological characterization of star-forming galaxies at $z>5$, and 4) to find $z>8$ galaxies magnified by the cluster lensing, with some bright enough to make them accessible to spectroscopic follow-up. Along with {\it HST}, the {\it Spitzer Space Telescope} has devoted 1000 hours of DD time to image the HFF fields at 3.6$\mu$m and 4.5$\mu$m with IRAC.

Whereas the main goal of the HFF is to explore the galaxy population in the first billion year of cosmic history, this dataset is also unique for its combination of surveyed area and depth for studies of galaxy evolution across most of the age of the universe, down to, and including, the redshifts of the targeted galaxy clusters ($z\approx0.3-0.5$).

The HFF data alone is however not sufficient to robustly characterize red galaxies at $z>3$ because the H$_{\rm 160}$ band lies on the UV side of the rest-frame optical Balmer/4000\AA~breaks at these redshifts, resulting in sub-optimal accuracies in the photometric redshifts and stellar population properties (e.g., stellar mass and rest-frame optical color) of galaxies at $z>2$ and even at $z>4$. Very deep K-band imaging is required to significantly improve the precision of both photometric redshifts and derived stellar population properties (see \S\ref{s:discussion}). Moreover, at $z>8-9$, the K-band data helps constraining the Lyman-break redshifts (e.g., \citealt{bouwens:13}) and increases the wavelength lever arm for measuring the redshift evolution of the rest-frame UV slopes (i.e., dust content and/or metallicity) of the first galaxies (\citealt{bouwens:12a}; \citealt{bouwens:13}).

To resolve this issue, i.e., the lack of K-band data over the HFF, we executed a program (dubbed {\it HFF-K$_s$} \textcolor{red}{\bf xxx how about finding a name for this program?}) to image all twelve HFF pointings in the $K_s$ band down to a comparable depth of the {\it HST} data using the instruments HAWK-I (mounted on the {\it VLT}) and MOSFIRE (mounted on the {\it Keck} telescope). This paper describes the {\it HFF-K$_s$} program and accompanies the data release. This paper is structured as follows. The observations and data reduction are presented in \S\ref{s:observations}, while \S\ref{s:mosaic} presents the drizzled K$_s$ mosaics, along with the quantification of the image quality, the noise properties and depth, and the completeness. Finally, \S\ref{s:discussion} concludes the paper showing a few crucial improvements introduced by the K$_s$ data.

@@

$K_s$-band observations of the Abell 2744 and MACS-0416 fields were obtained between 2013 Oct and 2014 Feb with the High Acuity Wide-field K-band Imager \citep[\HAWKI;][]{hawki, hawki2} on the 8.2\,m UT4 telescope at the ESO \textit{Very Large Telescope} (ESO program \mbox{092.A-0472}).  \HAWKI\ $K_s$-band observations of the Abell S1063 and Abell 370 fields were obtained between 2015 Jul and 2016 Jan from a subsequent program (095.A-0533).  \HAWKI\ is composed of four chips with 2048$\times$2048 $0\farcs106$ pixels, and the full $7\farcm5\times7\farcm5$ \HAWKI\ field-of-view is perfectly suited to cover both the primary and parallel \textit{HST} ACS optical and WFC3 IR fields simultaneously in a single pointing (see Fig.\ref{fig:layout}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   Exposure Map figure
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}
\epsscale{1.17}
\plotone{hawkiff_layout.pdf} % Figures/exposure_map.pdf
\caption{Layout of the Frontier Fields $K_s$-band mosaics.  The positions of the \textit{HST} cluster and parallel fields are shown in the blue (ACS optical) and red (WFC3 IR) polygons.  The light blue polygons in the MACS-0717 field show additional wide-field ACS imaging coverage from  programs GO-9722 and GO-10420 \citep[PI: Ebeling;][]{ma:11}.  The area covered by deep imaging in the \textit{Spitzer}/IRAC 3.6 and 4.5\,\mum\ channels is shown in orange, with the cluster fields indicated by the dashed lines.  The \HAWKI\ field of view is perfectly suited (Abell 2744, MACS-0416, Abell S1063 and Abell 370) for simultaneous imaging of the cluster+parallel field pairs, which require two separate pointings with MOSFIRE (MACS-0717 and MACS-1149).  The footprint of additional archival \HAWKI\ coverage of the MACS-1149 field is indicated by the gray square in the lower right panel.\label{fig:layout}}  
\end{figure*}
%%%%%%%%%%%%%%%%

@@

The \HAWKI\ and MOSFIRE observations were reduced with a pipeline that has been developed for previous surveys with the NEWFIRM \citep[NMBS;][]{whitaker:nmbs} and FOURSTAR (zFOURGE; \citealt{spitler:12}; Labb\'e et al., in prep.) infrared imaging instruments.  Treating each detector individually, the pipeline is easily modified for the different instrument configurations of the four \HAWKI\ and single MOSFIRE chips.  The primary task of the pipeline is removing the bright, time-variable sky background from the individual exposures, which is often some $10^4$ times brighter than the distant galaxies of interest in the field.  With such a bright background, we first determine an empirical ``sky flat'' that is a median of all of the \textit{science} exposures in a \HAWKI\ OB or MOSFIRE group, after rejecting the brightest 12 exposures at each pixel position to remove the contribution of bright objects.  We find these empirical flats to be preferable to external twilight or dome flats given the difficulty of obtaining a truly flat illumination pattern over such large detector fields-of-view.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   2MASS comparison
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!t]
\epsscale{1.17}
\plotone{2mass_phot.pdf} % Figures/exposure_map.pdf
\caption{Comparison of bright-star photometry to the 2MASS public catalog.  The photometric calibration of the $K_s$-band images were determined from standard star observations taken concurrently with the science exposures.  The uncertainties shown are taken from the 2MASS catalog; these stars are measured at very high S/N in the deep $K_s$-band mosaics.  We estimate the overall systematic uncertainty in the photometric calibration to be $\lesssim0.05~\mathrm{mag}$.\label{fig:2mass}}  
\end{figure}
%%%%%%%%%%%%%

@@

The 2-Micron All Sky Survey \citep[2MASS;][]{2mass} provides an additional check on the photometric calibration, though the comparison is limited due to relatively little brightness overlap between the faint end of reliable 2MASS photometry and the bright end where stars are in the linear regime of the detectors on the 8--10\,m telescopes.  A comparison of the observed photometry in the four survey fields to the 2MASS catalog magnitudes is shown in Figure~\ref{fig:2mass}.  The stellar photometry on the deep $K_s$-band mosaics is measured within $1\arcsec$ apertures corrected to infinity with the curves of growth described below.  The error-bars on the points in Figure~\ref{fig:2mass} come from the 2MASS catalog; the photometric uncertainties are negligible for such bright stars in the deep images ($\mathrm{S/N}>100$).  The 2MASS comparison suggests a systematic offset of $-0.09$~mag for the Abell S1063 field.  However, the exposure time for that field was similar to those of the additional deep \HAWKI\ fields and the derived depths of \textit{all} deep \HAWKI\ fields are nearly identical (Tab.~\ref{tab:summary}).  Given the overall agreement with the 2MASS photometry, we estimate that the systematic uncertainty of the absolute photometric calibration is $\sigma_\mathrm{sys}\leq0.05$~mag.  \textcolor{red}{\bf xxx It is not clear here if the offset of -0.09 mag found for Abell S1063 was actually applied to the K-band mosaic... Given your argument, I am guessing it was not, approach that I would agree with, but it is not clear... probably worth being explicitly... }

@@

In order to simplify measurements from the $K_s$-band images in consort with space-based \textit{HST} and \textit{Spitzer} mosaics, the astrometry of the ground-based mosaics must be refined.  Reference absolute astrometric catalogs were generated from \textit{HST} images when available (i.e., the chips overlapping the cluster and parallel \textit{HST} fields) and public \textit{Subaru} Suprime-Cam $r_c$-band images otherwise\footnote{\url{http://archive.stsci.edu/pub/hlsp/clash/}}.  Next, object catalogs are generated with the SEXtractor software \citep{bertin:96} for each individual background-subtracted exposure and transformations to the reference frame (shift, rotation and scale) are computed with the \texttt{stsci.stimage.xyxymatch} and \texttt{scikit-image.transform}\footnote{\url{http://scikit-image.org}} Python software tools.  For the \HAWKI\ exposures, we fit a third-order polynomial to the geometric distortion model determined by \cite{libralato:14} and specify the polynomial terms as ``Simple Imaging Polynomial'' (SIP) coefficients \citep{fits:sip} in the individual exposure FITS files.  The SIP FITS header generally applicable for \HAWKI\ is provided in Appendix \xxx.  We note that the \HAWKI\ distortion is generally small, reaching $\sim$2 pixels ($0\farcs2$) at the image corners, but that the distortion corrections are necessary, in particular, given the excellent overall image quality of the observations.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   Drizzle figure
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[!ht]
\epsscale{1.17}
\plotone{ivar_mosaics.pdf} % Figures/exposure_map.pdf
\caption{Inverse variance maps of the $K_s$-band mosaics.  The orientation of the fields is the same as that shown in Fig.~\ref{fig:layout}.  The four \HAWKI\ chips and two MOSFIRE pointings are clearly visible for the first four and last two fields, respectively.
\label{fig:ivar_mosaics}}  
\end{figure}
%%%%%%%%%%%%%%%%%%%

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%   Image quality figure
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure*}[!t]
\epsscale{1.17}
\plotone{curve_of_growth.pdf} % Figures/exposure_map.pdf
\caption{Curves of growth of stellar profiles in the Frontier Fields $K_s$-band mosaics.  The inverse aperture corrections for point sources in $D=0\farcs6$ apertures are indicated in each panel.  The profiles of stars are well-fit by Moffat profiles with $\mathrm{FWHM}\sim0\farcs4$ in most cases and $\beta\sim2.1$.  These extended Moffat profiles appear to be characteristic of deep $K_s$-band images and they have substantially more flux at large radii ($r>\mathrm{FWHM}/2$) than Gaussian profiles with the same FWHM (orange curves in the left panels). \textcolor{red}{\bf xxx In the bottom right panel, the inverse aperture correction of the parallel field is impossible to read, behind the other number...you may want to shift it upward a but}  \label{fig:cog}}  
\end{figure*}
%%%%%%%%%%%%%%

@@

Finally, we subtract a cell-based background from the final mosaics using an algorithm based on that used by SExtractor and SWarp but that provides more aggressive masking of flux in the outer isophotes of bright galaxies. The mosaics are shown in Figure~\ref{fig:layout}, along with the position of the deep Frontier Fields \textit{HST} imaging fields.  We also compute the robust NMAD scatter \citep{brammer:08} of empty sky pixels within the same cells to empirically calibrate the inverse variance maps that are shown in Figure~\ref{fig:ivar_mosaics} (see also \S\ref{s:depth}).  The final image and inverse variance mosaics are provided as an ESO Phase 3 data product \xxx\ URL, as well as attached as a digital product to this manuscript (\xxx\ I think ApJSS can and wants to do this).  The released images are all scaled to a common zeropoint of 26.0 (AB mag), which gives pixel values of order unity faint galaxies near the detection threshold.  In the sections below we describe the characteristics of the mosaics in more detail.
